---
title: "Statistical analysis of inter-event times for ASOIAF"
author: "Colm Connaughton"
date: "03/07/2020"
output:
  pdf_document: default
  html_document: default
---

We will use the poweRlaw library extensively:
```{r}
library(poweRlaw)
```
Set seed for reproducibility.
```{r}
set.seed(1)
```

Read in the data for the inter-event times for chapters containing significant deaths 
as measured both in discourse time (chapters) and story time (Westerosi days):

```{r}
data_folder <- "./data/4-analysis/"
filename <- paste(data_folder,"death_intervals_by_chapter.txt" ,  sep="")
data <- read.csv(filename, header=FALSE, sep=",")
d.discourse = data[['V1']]

filename <- paste(data_folder,"death_intervals_by_date.txt" ,  sep="")
data <- read.csv(filename, header=FALSE, sep=",")
d.story = data[['V1']]
```
## Geometric (discrete exponential) distribution

First we model the data by a discrete exponential (geometric) distribution and
estimate the best fit parameter using the method of maximum likelihood:

```{r}
# Create new distribution objects
m.exp.discourse = disexp$new(d.discourse)
m.exp.story = disexp(d.story)

# This is the parameter estimation bit
est.exp.discourse = estimate_pars(m.exp.discourse)
m.exp.discourse$setPars(est.exp.discourse$pars)

est.exp.story = estimate_pars(m.exp.story)
m.exp.story$setPars(est.exp.story$pars)

# Print out the most likely parameters
cat('Story time: geometric fit q = ', est.exp.story$pars)
cat('\nDiscourse time: geometric fit q = ', est.exp.discourse$pars)
```

Here are the plots of the best fit geometric models. We can already see a big difference.
```{r}
plot(m.exp.discourse)
lines(m.exp.discourse, col=2)
```

```{r}
plot(m.exp.story)
lines(m.exp.story, col=2)
```
We can estimate uncertainties in the parameter values and obtain goodness of fit p-values 
by bootstrapping the data. We will define statistical significance to correspond to a 
p-value threshold of 0.05 in line with a 95% confidence level.

Here we fix xmin=1 since we want to fit a pure geometric distribution rather than
allow thresholding of the data to get a better fit to the tail of the distribution.

```{r}
bs.exp.story = bootstrap_p(m.exp.story, no_of_sims=1000, xmin=1)
bs.exp.discourse = bootstrap_p(m.exp.discourse, no_of_sims=1000, xmin=1)
```
```{r}
plot(bs.exp.discourse)
```

Print out the best fit parameter value with approximate 95% confidence interval
```{r}
paramVal = mean(bs.exp.discourse$bootstraps[,3])
probs = c(2.5, 97.5)/100
ci = quantile(bs.exp.discourse$bootstraps[,3], probs)
pValue = bs.exp.discourse$p
cat("Discourse time: geometric (exponential) fit:")
cat("\n            q = ", paramVal, ', [', ci[[1]], ' , ', ci[[2]],']')
cat("\n            p-value = ", pValue )
```
From this p-value, we **fail to reject** the null hypothesis that the inter-event times for
significant deaths in discourse time follow a geometric distribution.

```{r}
plot(bs.exp.story)
```
```{r}
paramVal = mean(bs.exp.story$bootstraps[,3])
probs = c(2.5, 97.5)/100
ci = quantile(bs.exp.story$bootstraps[,3], probs)
#stddev = sd(bs.exp.story$bootstraps[,3])
#conf=2.0*stddev
pValue = bs.exp.story$p
#cat("Discourse time: q = ", paramVal, ' +/- ', conf)
#cat("\nDiscourse time: Goodness of fit p-value = ", pValue )
cat("Story time: geometric (exponential) fit:")
cat("\n            q = ", paramVal, ', [', ci[[1]], ' , ', ci[[2]],']')
cat("\n            p-value = ", pValue )
```


The p-value is not literally zero - just too small to report. From this p-value, we **reject** the null hypothesis 
that the inter-event times for significant deaths in story time follow a geometric distribution.

## Power law distribution (with lower cut-off)

Based on the analysis above, we conclude that the data on inter-event times for significant deaths in discourse
time is consistent with a geometric (discrete exponential) distribution but the corresponding data in story time is not. Let us now see if a power law distribution can better describe the story time data.

```{r}
m.pl.story = displ$new(d.story)
m.pl.discourse = displ$new(d.discourse)

# Estimate the x_min value for the lower cut-off and alpha value
xMin.story = estimate_xmin(m.pl.story)
m.pl.story$setXmin(xMin.story)

xMin.discourse = estimate_xmin(m.pl.discourse)
m.pl.discourse$setXmin(xMin.discourse)

cat('Story time: estimated xmin = ', xMin.story$xmin)
cat('\n            estimated alpha = ', xMin.story$pars)
cat('\nDiscourse time: estimated xmin = ', xMin.discourse$xmin)
cat('\n            estimated alpha = ', xMin.discourse$pars)


```

Make a plot of the fitted model
```{r}
plot(m.pl.story)
lines(m.pl.story, col=2)
```
```{r}
plot(m.pl.discourse)
lines(m.pl.discourse, col=2)
```


Perform bootstrapping to estimate uncertainties and goodness of fit for story time:

```{r}
bs.pl.story = bootstrap_p(m.pl.story, no_of_sims=1000)
plot(bs.pl.story)
```

```{r}
paramVal.alpha = mean(bs.pl.story$bootstraps[,3])
probs = c(2.5, 97.5)/100
alpha.ci = quantile(bs.pl.story$bootstraps[,3], probs)
paramVal.x0 = mean(bs.pl.story$bootstraps[,2])
x0.ci = quantile(bs.pl.story$bootstraps[,2], probs)
pValue = bs.pl.story$p
#cat("Story time: alpha = ", paramVal.alpha, ' +/- ', conf.alpha)
#cat("\n            x0    = ", paramVal.x0, ' +/- ', conf.x0)
#cat("\nStory time: Goodness of fit p-value = ", pValue )
cat("Story time: power law fit:")
cat("\n            alpha = ", paramVal.alpha, ', [', alpha.ci[[1]], ' , ', alpha.ci[[2]],']')
cat("\n            x0    = ", paramVal.x0, ', [', x0.ci[[1]], ' , ', x0.ci[[2]],']')
cat("\n            p-value = ", pValue )
```
From this p-value, we **fail to reject** the null hypothesis that the inter-event times for significant deaths in story time follow a power law distribution with exponent 2.0  and lower cut-off 3.5.
```{r}
bs.pl.discourse = bootstrap_p(m.pl.discourse, no_of_sims=1000)
plot(bs.pl.discourse)
```
Perform bootstrapping to estimate uncertainties and goodness of fit for discourse time:


```{r}
paramVal.alpha = mean(bs.pl.discourse$bootstraps[,3])
probs = c(2.5, 97.5)/100
alpha.ci = quantile(bs.pl.discourse$bootstraps[,3], probs)
paramVal.x0 = mean(bs.pl.discourse$bootstraps[,2])
x0.ci = quantile(bs.pl.discourse$bootstraps[,2], probs)
pValue = bs.pl.discourse$p
cat("Discourse time: power law fit:")
cat("\n            alpha = ", paramVal.alpha, ', [', alpha.ci[[1]], ' , ', alpha.ci[[2]],']')
cat("\n            x0    = ", paramVal.x0, ', [', x0.ci[[1]], ' , ', x0.ci[[2]],']')
cat("\n            p-value = ", pValue )
```
From this p-value, we also **fail to reject** the null hypothesis that the inter-event times for significant deaths in discourse time follow a power law distribution with exponent 3.9 and lower cut-off 3.7 but the uncertainties in these values are very large.


